---
title: "Creating a hotfix with git"
author: "Tom Jemmett"
date: "2023-03-24"
categories: [git, tutorial]
---

I recently discovered a bug in a code-base which needed to be fixed and deployed back to production A.S.A.P., but since the last release the code has moved on significantly. The history looks something a bit like:

```{r}
#! fig.height: 1
#! echo: false
library(ggplot2)

data.frame(
  x = 1:6,
  y = c(1, 1, 1, 1, 2, 2),
  g = c(1, 1, 1, 1, 1, 1),
  t = c("tag", "main", "main", "main", "branch", "branch"),
  l = c("v0.2.0", "main", "main", "main", "branch", "branch")
) |>
  ggplot(aes(x, y, group = g)) +
  geom_line() +
  geom_label(aes(label = l, fill = t)) +
  theme_void() +
  theme(legend.position = "none")
```

That is, we have a tag which is the code that is currently in production (which we need to patch), a number of commits after that tag to main (which were separate branches merged via pull requests), and a current development branch.

I need to somehow: 1. go back to the tagged release, 2. check that code out, 3. patch that code, 4. commit this change, but insert the commit before all of the new commits after the tag

There are at least two ways that I know to do this, one would be with an [interactive rebase](https://about.gitlab.com/blog/2020/11/23/keep-git-history-clean-with-interactive-rebase/), but I used a slightly longer method, but one I feel is a little less likely to get wrong.

Below are the step's that I took. One thing I should note is this worked well for my particular issue because the change didn't cause any merge conflicts later on.

## Fixing my codebase

First, we need to checkout the tag

```         
git checkout -b hotfix v0.2.0
```

This creates a new branch called `hotfix` off of the tag `v0.2.0`.

Now that I have the code base checked out at the point I need to fix, I can make the change that is needed, and commit the change

```         
git add [FILENAME]
git commit -m "fixes the code"
```

*(Obviously, I used the actual file name and gave a better commit message. I Promise ðŸ˜)*

Now my code is fixed, I create a new tag for this "release", as well as push the code to production (this step is omitted here)

```         
git tag v0.2.1 -m "version 0.2.0"
```

At this point, our history looks something like

```{r}
#! fig.height: 1
#! echo: false
data.frame(
  x = c(1,2,2:6),
  y = c(1, 2, 1, 1, 1, 2, 2),
  g = c(1, 1, 2, 2, 2, 2, 2),
  t = c("tag", "tag", "main", "main", "main", "branch", "branch"),
  l = c("v0.2.0", "v0.2.1", "main", "main", "main", "branch", "branch")
) |>
  ggplot(aes(x, y, group = g)) +
  geom_line() +
  geom_line(
    data = data.frame(x = c(1, 2), y = c(1, 1), g = c(1, 1))
  ) +
  geom_label(aes(label = l, fill = t)) +
  theme_void() +
  theme(legend.position = "none")
```

What we want to do is break the link between main and `v0.2.0`, instead attaching to`v0.2.1`. First though, I want to make sure that if I make a mistake, I'm not making it on the main branch.

```         
git checkout main
git checkout -b apply-hotfix
```

Then we can fix our history using the rebase command

```         
git rebase hotfix
```

What this does is it rolls back to the point where the branch that we are rebasing (`apply-hotfix`) and the `hotfix` branch both share a common commit (`v0.2.0` tag). It then applies the commits in the `hotfix` branch, before reapplying the commits from `apply-hotfix` (a.k.a. the `main` branch).

One thing to note, if you have any merge conflicts created by your fix, then the rebase will stop and ask you to fix the merge conflicts. There is some information in the GitHub doc's for \[resolving merge conflicts after a Git rebase\]\[2\].

\[2\]: https://docs.github.com/en/get-started/using-git/resolving-merge-conflicts-after-a-git-rebase

At this point, we can check that the commit history looks correct

```         
git log v0.2.0..HEAD
```

If we are happy, then we can apply this to the `main` branch. I do this by renaming the `apply-hotfix` branch as `main`. First, you have to delete the `main` branch to allow us to rename the branch.

```         
git branch -D main
git branch -m main
```

We also need to update the other branches to use the new main branch

```         
git checkout branch
git rebase main
```

Now, we should have a history like

```{r}
#! fig.height: 1
#! echo: false
data.frame(
  x = 1:7,
  y = c(1, 1, 1, 1, 1, 2, 2),
  g = c(1, 1, 1, 1, 1, 1, 1),
  t = c("tag", "tag", "main", "main", "main", "branch", "branch"),
  l = c("v0.2.0", "v0.2.1", "main", "main", "main", "branch", "branch")
) |>
  ggplot(aes(x, y, group = g)) +
  geom_line() +
  geom_label(aes(label = l, fill = t)) +
  theme_void() +
  theme(legend.position = "none")
```

## 
